Extracting Function Signature from R Program Traces

### 2020-05-13 paper/main.tex
reverse dependency eval disc

Functions are first-class values in R.

\typetracer found \AT{number} functions to be higher-order, having at least one function type among their arguments.
They are markedly more common in a subset of packages, such as \AT{such as...}.
By-and-large, these higher-order functions are similar to well-known higher-order functions like map, filter, and fold.

@@

We contended with multiple type language designs, and in this section we will explore in more detail how we compared our candidate designs.
We will also discuss a large-scale evaluation of our final set of type annotations.

%
%
%
%
\subsection{Comparative Analyses}

\AT{We compared different designs: with/out structs and tuples, with/out classes, with/out primitive subtyping. We looked at type sizes and determined that e.g. \code{? T} is a good type.}

%
%
\subsubsection{Analysis of Polymorphism}

\AT{How many arguments, functions are polymorphic? How many packages? Are there pathologically polymorphic packages? Why didn't you handle those?}

@@

\AT{When is any introduced? Why? How does it affect the results? What if missing/unused isn't any, do the results change substantially?}

%
%
%
%
\subsection{Large-Scale Evaluation}

Now that we have settled on a language design, we endeavour to evaluate it against a representative subset of existing R code.
To this end, we took the detailed type information we collected from \AT{corpus} popular R packages, inferred types for the functions exposed by those packages, and ran the test, example, and vignette code of each of their reverse dependencies.

%
%
\subsubsection{Experiment Setup}

\AT{What is the design of the experiment? The setup? How much code was ran? What does this evaluate? Why does it mean anything for the annotation framework?}

\typetracer is capable of inferring type signatures based on the call traces it has collected.
These call traces were consolidated into a single function type for each package function:
\typetracer employs unification strategies discussed in Section~\ref{sec:typesystemdesign} to keep the size of signatures in check and only generates an \code{any} type under extenuating circumstances: when inner list types got too big (we decided to cap the union type in lists to 5), when missing and unused arguments were recored, and finally when a longjump caused function execution to abort prematurely.
Finally, we loaded each of these inferred types into \contractr, and ran all example code from the reverse dependencies of these popular packages.

@@

Overall, we found that only \AT{2\%} of contract assertions failed.
\AT{What does this indicate? One thought:}
This indicates two things: firstly, the corpus of packages we selected is quite well-tested, and much of the intended function behaviour is exemplified in the package example code. 
Secondly, and more importantly, this success indicates that our analysis framework is able to generate function types which accurately capture the behaviour of the function. 
\AT{It kind of feels like these need to simultaneously be true. Maybe we pull in the fact that coverage is good?}

Of the \AT{2\%} of failing contract assertions, we picked up on a few significant patterns.
Firstly, we found that \AT{10\%} of the failing assertions were due to \code{double}-typed values being passed to function arguments expecting an \code{integer}.
While we cannot say for sure that {\it all} of these cases could be handled by coercion, it is worth noting that coercion from doubles to integers is exceedingly common in R: the practice is not codified in the language spec, but many of R's internal C functions explicitly perform this coercion.
\AT{We can put an example, e.g. nrow(df) is an integer but 5 is not. Also list access coerces the double to an int.}

@@

Separately, we investigated the effect of running reverse dependency code on the code coverage of the packages in our corpus.
We found that \AT{in many cases}. the reverse dependencies increased the code coverage (this aligns with findings in \AT{cite genthat}), and this was especially true for the functions for which contract assertions failed.

>>>>

\AT{Revisit this section since discussion of polymorphism in results.}

Functions are first-class values in R.
\typetracer found \AT{number} functions to be higher-order, having at least one function type among their arguments.
They are markedly more common in a subset of packages, such as \AT{such as...}.
By-and-large, these higher-order functions are similar to well-known higher-order functions like map, filter, and fold.

@@

To evaluate the effectiveness of our type annotation language and contract checking framework, we conducted a large-scale experiment based on the corpus of 412 packages we used to inform our language design.
This corpus represents some of the most widely used and thoroughly tested R libraries on CRAN.

We configured \typetracer to infer the types of each of the functions that are visible to package clients.
Recall that \typetracer's primary function is to collect call signatures from observed function calls, and we ran \typetracer on the test, example, and vignette code of all packages in the aforementioned corpus.
These call traces were consolidated into a single function type for each visible package function:
\typetracer employs unification strategies discussed in Section~\ref{sec:typesystemdesign} to keep the size of signatures in check and only generates an \code{any} type when missing and unused arguments were recored, and when a longjump caused function execution to abort prematurely.
The types generated by \typetracer will be analyzed in more depth in Section~\ref{subsec:stats}, and a large-scale evaluation of the quality of these types will be discussed in Section~\ref{subsec:eval}.

%
%
%
%
\subsection{Language Statistics}
\label{subsec:stats}

We analyzed the type signatures generated by \typetracer in order to build up a profile of our type annotation language.
We were interested in broad, high-level information, such as the distribution of generated types, the level of polymorphism observed in the types, and the distribution of sizes of union types.

%
%
\subsubsection{Type Distribution}

%
%
\subsubsection{Polymorphism}

%
%
\subsubsection{Size of Union Types}

@@

%
%
\subsubsection{Select Examples}

%
%
%
%
\subsection{Large-Scale Evaluation}
\label{subsec:eval}

To measure the effectiveness of our type annotation language, we designed an experiment pitting the types generated by a package's own testing code against the testing code of the package's clients.
We collected failed contract assertions in order to analyze them and perhaps identify weaknesses in our type annotation language.

%
%
\subsubsection{Experimental Design}

As mentioned earlier in this section, we configured \typetracer to infer types for all client-visible functions for each of the 412 packages in our corpus.
We loaded each of these inferred types into \contractr, and ran all example code from the reverse dependencies of these popular packages.
\AT{As discussed in genthat}, running R code can be tricky, as runnable code may depend on e.g. data present in a hard-coded directory structure, and we employed similar strategies to ensure that we ran as much code as possible.
\AT{Maybe Filip can write a few sentences on this, or write me a message and I can work it in.}

\contractr was equipped to provide a list of all failed contract assertions, complete with diagnostic information (including a stack trace) to help us make sense of the failed assertion.

@@

Overall, we found that only \AT{2\%} of contract assertions failed.
This indicates two things: firstly, the corpus of packages we selected is quite well-tested, and much of the intended function behaviour is exemplified in the package example code. 
Secondly, and more importantly, this success indicates that our analysis framework is able to generate function types which accurately capture the behaviour of the function. 
Of the \AT{2\%} of failing contract assertions, we picked up on a few significant patterns.

Firstly, we found that \AT{10\%} of the failing assertions were due to \code{double}-typed values being passed to function arguments expecting an \code{integer}.
While we cannot say for sure that {\it all} of these cases could be handled by coercion, it is worth noting that coercion from doubles to integers is exceedingly common in R: the practice is not codified in the language spec, but many of R's internal C functions explicitly perform this coercion.
\AT{We can put an example, e.g. nrow(df) is an integer but 5 is not. Also list access coerces the double to an int.}

@@

Of the remaining \~60\% of failed contract assertions, we found quite an interesting pattern: the top as-of-yet unexplained contract failure, making up roughly 7\% of total failed assertions, is an assertion checking a ${\bf character}$ against the expected type ${\bf class} \langle bignum \rangle \, | \, {\bf raw}[]$.

Separately, we investigated the effect of running reverse dependency code on the code coverage of the packages in our corpus.
Further investigation revealed that most failures were coming from the \code{openssl::base64_encode} function, whose code is available in Figure~\ref{fig:base64-encode}.

\begin{lstlisting}
# The type according to typetracer:
#' @type <bin: class<`bignum`> | raw[], linebreaks: logical> => character
base64_encode <- function (bin, linebreaks = FALSE) 
{
    if (is.character(bin)) {
        bin <- charToRaw(paste(bin, collapse = ""))
    }
    stopifnot(is.raw(bin))
    .Call(R_base64_encode, bin, as.logical(linebreaks))
}
\end{lstlisting}

\caption{Example of reverse dependency increasing code coverage.}
\label{fig:base64-encode}
\end{center}
\end{figure}

Contract failures are on the first argument, \code{bin}.
This function is a wrapper for a C function (\code{R_base64_encode}) which encodes the input \code{bin} in base64.
It would appear that the code for this function was never tested with a ${\bf character}$-typed input, even though it {\it is} valid.
This is an example of a rather common phenomenon wherein the code of a package's reverse dependencies increases the coverage of some of the original package's functions, exemplified here by the code on line 6 never having been run in \code{openssl}'s own tests.

We found that \AT{in many cases}. the reverse dependencies increased the code coverage (this aligns with findings in \AT{cite genthat}), and this was especially true for the functions for which contract assertions failed.
